Global code motion of congruent instructions

### 2017-04-18 paper.tex
page

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a \gcm{} (\GCM{}) compiler optimization which schedules
congruent instructions across the program.  Not only this saves code size, it
exposes redundancies in some cases, it exposes more instruction level
parallelism in the basic-block to which instructions are moved, and it enables
other passes like loop invariant motion to remove more redundancies. The cost
model to drive the code motion is based on live range analysis on \SSA{}
representation such that the (virtual) register pressure does not increase, which
translates to 2\% fewer spills while compiling \SPEC{} benchmark suite.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on overall compile time on \SPEC{}. \GCM{} enables more inlining
and exposes more loop invariant code motion opportunities in majority of the
benchmarks. We have also seen execution time improvements in a few of \SPEC{}
benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
\SPEC{} benchmarks suite when compiled for \xlinux{}. This is an optimistic
algorithm in the sense that it considers all congruent instructions in a
function as potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case a common hoisting point for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of \gcm{}, and how
\gcm{} increases instruction level parallelism thereby enabling more \ooo{}
execution on modern architectures.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like CSE \cite{dragonbook} work very well on single basic-blocks but
when it comes to detect redundancies across basic-blocks these techniques fall
short: more complex passes like GCSE and \PRE{} have been designed to handle these
cases based on data-flow analysis \cite{morel1979global}.  At first these
techniques were described in the classical data-flow analysis framework, and
later the use of the \SSA{} representation lowered the cost in terms of compilation
time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
techniques in the main stream: nowadays \SSA{} based \PRE{} is available in every
industrial compiler.

This paper describes \gcm{} (\GCM{}) of congruent instructions \cite{briggs1997}, a
technique that uses the information computed for \PRE{} to detect identical
computations but has a transformation phase whose goal differs from \PRE{}: it
removes identical computations from different branches of execution.  These
identical computations in different branches of execution are not redundant
computations at run-time and the number of run-time computations is not
reduced. It is not a redundancy elimination pass, and thus it has different cost
function and heuristics than \PRE{} or CSE. It is more similar to global code
scheduling \cite{dragonbook,click1995global} in the sense that it will only move
computations. Code hoisting, for example, can reduce the critical path length of
execution in \ooo{} machines. As more instructions are available at the
hoisting point, the hardware has more independent instructions to reorder. The
following example illustrates how hoisting can improve performance by exposing
more ILP.

@@

\begin{verbatim}
// Return true if the merge-set of source node of a visited J-edge (incoming edge of
// lnode) is not the subset of the merge-set of lnode.
bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet MergeSet)
  for e in (all incoming edges to lnode)
    if e is in JE and e in V
      if MergeSet(Source Node of e) is not a subset of MergeSet(lnode)
        return true
  return false

// Compute a merge-set top-down in breadth first order.
bool constructMergeSet_1(BFSList B, JEdges JE, DomLevel DL)
  repeat = false
  V = empty list of visited edges
  for n in B
    for e in (all incoming edges to n)
      if e is in JE and e not in V
        V(e) = true
        let tmp = Source Node of e
        let tnode = Target Node of e
        let lnode = NULL
        while (DL(tmp) >= DL(tnode))
          MergeSet(tmp) = MergeSet(tmp) U MergeSet(tnode) U {tnode}
          lnode = tmp
          tmp = dom-parent(tmp)
        repeat = still_inconsistent(lnode, JE, MergeSet)
  return repeat

// Construct merge-set of each node in Control-Flow-Graph G of a function.
void constructMergeSet(CFG G)
  B = Breadth First Order of G
  JE = JEdges of G
  DL = List of Path length (from root) of each node in G.
  do // Call until a fixed point is reached.
    Repeat = constructMergeSet_1(B, JE, DL)
  while (Repeat)
\end{verbatim}

@@

  // Iterate over all the uses of A and see if any intersect with the merge-set of N.
  for T in users(A)
    B = basic_block(T)
    while (B != null and B != def_bb(A))
      if B in MergeSet(N) return true
      B = dom-parent(B)
  return false

// Return true if op is the last use of the operand in I.
bool isKill(Operand op, Instruction I)
  if (isLiveOutUsingMergeSet(op))
    return false

  for each I1 after I in basic_block(I)
    if (I1 uses op)
      return false
  return true
\end{verbatim}

The original algorithm presented in \cite{das2012} has mismatched types in terms
of uses and nodes because each node can have many instructions and hence many
uses. Also, while iterating on the dominator of each user of a variable we may
reach to the beginning of a function, in that case the inner while loop needs to
terminate. These two cases were missing from the algorithm and we came across
them during implementation. If a variable is not live-out of a basic-block, it
still may be used later in the basic-block: to establish whether a use of a
variable is a kill we iterate on all the subsequent instructions in a
basic-block checking for uses, the \emph{isKill} function uses
\emph{isLiveOutUsingMergeSet} to find out if an operand is killed in an
instruction.

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instructions before and after its motion
will produce the same side-effects.  For hoisting in particular, we need to
prove that on all execution paths from the hoisting point to the end of the
program the side effects appear in the same order.

\begin{verbatim}
// Compute downward safety of hoistable instructions at hoist-point
void compute_downward_safety()
  Worklist = list of all basic-blocks of hoistable instructions
  for each basic-block B in the dominator tree starting at hoist-point:
    if Worklist is empty, return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      remove subtree with root at B // Available at B => downward safe from B
    if B is a leaf basic-block, return false  // Path exists! not downward safe
    if B dominates hoist-point, return false  // Back edge
  return true
\end{verbatim}

@@

\begin{verbatim}
void doGCM()
  Analyses available: Dominator Tree, DFS Numbering, Memory SSA,
  Compute DJ-graph of function and compute mergeset based on that
  do
    Compute GVN of each expression in the function
  while (doCodeHoisting() > 0) // Repeat hoisting if any candidates were hoisted.
  doCodeSinking()
\end{verbatim}

We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
sinking. So first we find the common dominator dominating all such identical
computations and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithm would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators. However,
it should be noted that we only hoist very busy expressions to avoid checking
for undefined behaviors resulting because of introducing extra computation in an
execution path.

\begin{verbatim}
int doCodeHoisting() // Find hoistable candidates and do hoisting
  For each value number VN with 2 or more instructions (I1, I2, ..., Ik)
    sort the instructions according to DFSIn numbers
    if first two I1, I2 are hoistable and if they are downward-safe at hoist-point
          and if they are profitable to be hoisted at hoist-point
      proceed to check the ability to hoist subsequent instruction with same VN
    else
      proceed to check if I2 and I3 (if available) can be hoisted
  For each set of instructions (I1, ..., Ik) with same VN that can be hoisted:
    move the first one I1 to the hoist-point
    update the use of all other candidates (I2, ..., Ik) to refer to I1
    remove all other instructions (I2, ..., Ik)
    if I1 has memory references (a load, store, call, etc.)
      update MemorySSA of I1 and others to point to the MemorySSA reference of I1
      remove MemorySSA reference of all others which were deleted
    update statistics
  return number of hoisted instructions
\end{verbatim}

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a \gcm{} (\GCM{}) compiler optimization which schedules congruent
instructions across the program.  Not only \GCM{} saves code size, it exposes
redundancies in some cases, it exposes more instruction level parallelism in the
basic-block to which instructions are moved, and it enables other passes like
loop invariant motion to remove more redundancies. The cost model to drive the
code motion is based on liveness analysis on \SSA{} representation such that the
(virtual) register pressure does not increase resulting in 2\% fewer spills on
the \SPEC{} benchmark suite.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
inlining and exposes more loop invariant code motion opportunities in majority
of the benchmarks. We have also seen execution time improvements in a few of
\SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\%
on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
optimistic algorithm in the sense that it considers all congruent instructions
in a function as potential candidates. We make an extra effort to hoist
candidates by partitioning the potential candidates in a way to enable partial
hoisting in case a common hoisting point for all the candidates cannot be
found. We also formalize why register pressure reduces as a result of \gcm{},
and how \gcm{} increases instruction level parallelism thereby enabling more
\ooo{} execution on modern architectures.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like Common Subexpression Elimination (CSE) \cite{dragonbook} work
very well on single basic-blocks (local level) but when it comes to detect
redundancies across basic-blocks (global level) these techniques fall short:
more complex passes like Global Common Subexpression Elimination (GCSE) and
Partial Redundancy Elimination (\PRE{}) have been designed to handle these cases
based on data-flow analysis \cite{morel1979global}.  At first these techniques
were described in the classical data-flow analysis framework, but later the use
of Static Single Assignment (\SSA{}) representation lowered their cost in terms
of compilation time \cite{briggs1994effective,chow1997new,kennedy1999partial}
and brought these techniques in the main stream: nowadays \SSA{} based \PRE{} is
available in industrial compilers like \GCC{} and \LLVM{}.

This paper describes \gcm{} (\GCM{}) of congruent instructions
\cite{briggs1997}, a technique that uses the information computed for \PRE{} to
detect identical computations but has a transformation phase whose goal differs
from \PRE{}: it removes identical computations from different branches of
execution.  These identical computations in different branches of execution are
not redundant computations at run-time and the number of run-time computations
is not reduced. It is not a redundancy elimination pass, and thus it has
different cost function and heuristics than \PRE{} or CSE. It is similar to
global code scheduling \cite{dragonbook,click1995global} in the sense that it
will only move computations. Code hoisting, for example, can reduce the critical
path length of execution in \ooo{} machines. As more instructions are available
at the hoisting point, the hardware has more independent instructions to
reorder. The following example illustrates how hoisting can improve performance
by exposing more ILP.

@@

\begin{verbatim}
// Return true if the merge-set of source node of a visited J-edge (incoming edge of
// lnode) is not the subset of the merge-set of lnode.
bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet MergeSet)
  for e in (all incoming edges to lnode)
    if e is in JE and e in V
      if MergeSet(Source Node of e) is not a subset of MergeSet(lnode)
        return true
  return false
\end{verbatim}

Each basic block which has a join-edge (J-Edge) to its successor will have a
merge-set that is a subset of the merge-set of its successor \cite{das2005}. The
function \emph{still\_inconsistent} returns true if this invariant is not
satisfied which allows the function \emph{constructMergeSet} to repeat.

\begin{verbatim}
// Compute a merge-set top-down in breadth first order.
bool constructMergeSet_1(BFSList B, JEdges JE, DomLevel DL)
  repeat = false
  V = empty list of visited edges
  for n in B
    for e in (all incoming edges to n)
      if e is in JE and e not in V
        V(e) = true
        let tmp = Source Node of e
        let tnode = Target Node of e
        let lnode = NULL
        while (DL(tmp) >= DL(tnode))
          MergeSet(tmp) = MergeSet(tmp) U MergeSet(tnode) U {tnode}
          lnode = tmp
          tmp = dom-parent(tmp)
        repeat = still_inconsistent(lnode, JE, MergeSet)
  return repeat
\end{verbatim}
\newpage
\begin{verbatim}
// Construct merge-set of each node in Control-Flow-Graph G of a function.
void constructMergeSet(CFG G)
  B = Breadth First Order of G
  JE = JEdges of G
  DL = List of Path length (from root) of each node in G.
  do // Call until a fixed point is reached.
    Repeat = constructMergeSet_1(B, JE, DL)
  while (Repeat)
\end{verbatim}

@@

  // Iterate over all the uses of A and see if any intersect with the merge-set of N.
  for T in users(A)
    B = basic_block(T)
    while (B != null and B != def_bb(A))
      if B in MergeSet(N) return true
      B = dom-parent(B)
  return false
\end{verbatim}

If a variable is not live-out of a basic-block, it still may be used later in
the basic-block. To establish whether a use of a variable is a kill we iterate
on all the subsequent instructions in a basic-block checking for uses, the
\emph{isKill} function uses \emph{isLiveOutUsingMergeSet} to find out if an
operand is killed in an instruction.

\begin{verbatim}
// Return true if op is the last use of the operand in I.
bool isKill(Operand op, Instruction I)
  if (isLiveOutUsingMergeSet(op))
    return false
  for each I1 after I in basic block of I
    if (I1 uses op)
      return false // Not a kill
  return true
\end{verbatim}

The original algorithm presented in \cite{das2012} has mismatched types in terms
of uses and nodes because each node can have many instructions and hence many
uses. Also, while iterating on the dominator of each user of a variable we may
reach to the beginning of a function, in that case the inner while loop needs to
terminate. These two cases were missing from the algorithm and we came across
them during implementation.

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instruction before and after its
code-motion will produce the same side-effects.  For hoisting in particular, we
need to prove that on all execution paths from the hoisting point to the end of
the program the side effects appear in the same order. In general, we cannot
introduce a new computation along any path of execution without checking for
undefined behaviors. The algorithm \emph{compute\_downward\_safety} computes if
hoisting a set of instruction to a common hoisting point is downward-safe
\cite{steven1997advanced}. Since \GCM{} is only dealing with VBEs, computing
downward-safety ensures that no undefined behaviors are introduced as a result
of code-motion.
\newline
\begin{verbatim}
// Return true if a hoistable instruction is downward safe at hoist-point
bool compute_downward_safety()
  Worklist = list of all basic-blocks of hoistable instructions
  for each basic-block B in the dominator tree starting at hoist-point:
    if Worklist is empty
       return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      remove subtree with root at B // Available at B => downward safe from B
    if B is a leaf basic-block
      return false  // Path exists! not downward safe
    if B dominates hoist-point
      return false  // Back edge
  return true
\end{verbatim}

@@

\begin{verbatim}
void doGCM()
  Analyses available: Dominator Tree, DFS Numbering, Memory SSA,
  Compute DJ-graph of function
  compute mergeset based on that (using constructMergeSet function)
  compute barriers (using compute_barriers function)
  do
    Compute GVN of each expression in the function
  while (doCodeHoisting() > 0) // Repeat hoisting if any candidates were hoisted.
  doCodeSinking()
\end{verbatim}

We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm \emph{doGCM}
prefers hoisting (\emph{doCodeHoisting}) to sinking (\emph{doCodeSinking}).

For hoisting, we find a common dominator dominating all a set of congruent
instructions and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithm would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators.
\newpage
\begin{verbatim}
int doCodeHoisting() // Find hoistable candidates and do hoisting
  For each value number VN with 2 or more instructions (I1, I2, ..., Ik)
    sort the instructions according to DFSIn numbers
    if first two I1, I2 are hoistable and if they are downward-safe at hoist-point
          and if they are profitable to be hoisted at hoist-point
      proceed to check the ability to hoist subsequent instruction with same VN
    else
      proceed to check if I2 and I3 (if available) can be hoisted
  For each set of instructions (I1, ..., Ik) with same VN that can be hoisted:
    move the first one I1 to the hoist-point
    update the use of all other candidates (I2, ..., Ik) to refer to I1
    remove all other instructions (I2, ..., Ik)
    if I1 has memory references (a load, store, call, etc.)
      update MemorySSA of I1 and others to point to the MemorySSA reference of I1
      remove MemorySSA reference of all others which were deleted
    update statistics
  return number of hoisted instructions
\end{verbatim}
